Pyro -Velocity: Probabilistic RNA Velocity inference from single-cell data

### 2024-08-20 reproducibility/manuscript/v2.tex
chore(manuscript): update v2 diff file
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else  
    % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\setcounter{secnumdepth}{5}
% Make \paragraph and \subparagraph free-standing
\makeatletter
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}{
    \@ifstar
      \xxxParagraphStar
      \xxxParagraphNoStar
  }
  \newcommand{\xxxParagraphStar}[1]{\oldparagraph*{#1}\mbox{}}
  \newcommand{\xxxParagraphNoStar}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}{
    \@ifstar
      \xxxSubParagraphStar
      \xxxSubParagraphNoStar
  }
  \newcommand{\xxxSubParagraphStar}[1]{\oldsubparagraph*{#1}\mbox{}}
  \newcommand{\xxxSubParagraphNoStar}[1]{\oldsubparagraph{#1}\mbox{}}
\fi
\makeatother

@@

\raggedbottom
\usepackage{nameref}
\usepackage{placeins}
\allowdisplaybreaks
\makeatletter
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{plain}
\@ifundefined{c@chapter}{\newfloat{suppfig}{h}{losuppfig}}{\newfloat{suppfig}{h}{losuppfig}[chapter]}
\floatname{suppfig}{Supplementary Figure}
\newcommand*\listofsuppfigs{\listof{suppfig}{List of Supplementary Figures}}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother

@@

\section{Main text}\label{sec-main}

@@

Pyro -Velocity recasts the velocity estimation problem into a latent
variable posterior probability inference (Sec.~\ref{sec-methods}). The
proposed model is generative and fully Bayesian, with the different
parameters considered as latent random variables. Each parameter is
assigned a prior distribution, and the model is then conditioned on the
raw data (both spliced and unspliced molecule raw counts) to estimate
posterior probability distributions and to enable uncertainty estimation
of velocity. Central to the Pyro -Velocity model is a \emph{shared
latent time} \((t)\): a global latent variable placing each cell at a
particular instant of the modeled dynamical process. From this latent
time, the expectation of spliced and unspliced RNA levels for each gene
can be computed knowing gene-specific kinetics parameters
(transcriptional \((\alpha)\), splicing \((\beta)\), and degradation
rates (\(\gamma\)) and two gene-specific switching times
\((t_{0}^{k})\), and gene-specific transcriptional state \(k\)) and
integrating the RNA velocity differential equations analogously to what
proposed in the dynamical RNA velocity model by Bergen et al.
\citep{Bergen2020-pj, Li2021-qa}. However, we innovate the structure and
formulation of the latent time given that our model is set to jointly
fit the entire raw data matrix, not individual genes. We approach the
task of estimating the posterior distribution by restricting the search
within a family of a variational distribution approximating the true
posterior, and taking advantage of the automatic differentiation
variational inference (ADVI) capabilities of Pyro
\citep{Bingham2018-id, Kucukelbir2016-bk} (Fig.~\ref{fig-model}a,
Sec.~\ref{sec-methods}). Notably, our generative Bayesian model enables
to sample from the posterior distributions of the latent and dependent
variables, allowing uncertainty estimation of phase portraits, high
dimensional velocity vector field, and shared latent time. Around this
model, we built several visualizations to assess uncertainty and
functions to recover putative cell fate priming genes and to investigate
how the denoised gene expression recapitulates known cell-type
annotations and their developmental order (Fig.~\ref{fig-model}b,
Sec.~\ref{sec-methods}). Although several RNA velocity methods
\citep{Gayoso2024-fz, Gorin2022-il, Gu2022-rh, Cui2022-em, Gao2022-jq}
have been recently proposed, and many of them have not been
peer-reviewed, here we showcase the advantages and innovations of
Pyro -Velocity by comparing it with the current state-of-the-art
dynamical RNA velocity model proposed by Bergen et al.
\citep{Bergen2020-pj}, and implemented in scVelo.

@@

Pyro -Velocity was also able to define well-known developmental cell
hierarchies in a scRNA-seq dataset of E15.5 mouse pancreas
(Fig.~\ref{fig-model-results}b-e, \(n=3,696\) cells comprising \(8\)
cell types (Fig.~\ref{fig-model-results}b, panel 1)), identifying cell
trajectories originating from ductal progenitor cells and culminated in
the production of mature Alpha, Beta, Delta, and Epsilon cells
\citep{Bastidas-Ponce2019-lf}. Though the scVelo and Pyro -Velocity
models have qualitatively comparable vector fields, the results
generated by scVelo were greatly influenced by gene selection,
neighboring cell number for gene expression smoothing, and cannot
provide confidence assessment of lineage directionality
(Supplementary Figure~\ref{suppfig-model-results}a). In addition,
Pyro -Velocity uncertainty visualizations uncovered that differentiated
cells had a larger vector field uncertainty than stem and progenitor
cells, as would be expected for terminally differentiated cell states
where progression and lineage trajectories would not be anticipated.
Pyro -Velocity also recovered a developmental order in close agreement
with what was estimated by Cytotrace (Spearman's
\(r=0.95, \; \text{p-value} <
10^{-10}\) between latent time and developmental
order)\citep{Gulati2020-xq} - a state-of-the-art orthogonal method used
to predict cell differentiation based on the number of expressed genes
per cell.

We next used Pyro -Velocity to unbiasedly uncover potential cell fate
determinant markers genes within the E15.5 mouse pancreas. To this end,
we first selected the top 300 genes with the best goodness of fit
(negative mean absolute errors across cells). Next, we ranked them based
on the positive (or negative) correlation between inferred shared time
and corresponding denoised spliced expression per cell
(Fig.~\ref{fig-model-results}d, Methods). This procedure successfully
recovered key developmental genes that are activated (or repressed)
during endocrinogenesis (Fig.~\ref{fig-model-results}d-e,
Supplementary Figure~\ref{suppfig-model-results}). Among the top
positively correlated genes identified by Pyro -Velocity, endocrine Beta
cells expressed insulin 1 and 2 (Ins1, Ins2) and Nnat, which is a key
regulator Beta cell insulin generation \citep{Millership2018-vd} and
Ppp1r1a which is a marker of Beta cells \citep{Jiang2013-pt}. The
ranking procedure proposed by scVelo failed to identify Ins1, Ins2,
Nnat, and Ppp1r1a as lineage-restricted genes in their top 50 ranked
gene list. The inability of scVelo to identify these key developmental
genes could be likely explained by the kNN smoothing of read count and
filtering steps (Supplementary Table 1, 2). Visual inspection of these
four top genes by Pyro -Velocity revealed a single activation phase
during cell type transitions based on phase portrait curves. We also
propose their visualization through ``rainbow plots'\,' to illustrate
denoised single gene spliced expression across shared time and to
investigate the inferred cell order based on known annotations. Further,
UMAP embedding of cells colored based on denoised gene expression
confirmed the cell fate markers' smooth and progressive activation (or
repression) towards endocrine cells (Fig.~\ref{fig-model-results}e,
Supplementary Figure~\ref{suppfig-model-results}).

Ground truth velocity is not a quantity we can easily assess. Therefore,
no robust and quantitative benchmarking of RNA velocity methods has been
proposed. However, recent single-cell assays allow recording lineage
relationships and can be used to assess the ability of different methods
in recapitulating cell fate progression. Therefore, we next
systematically compared scVelo and Pyro -Velocity ability to recover
single cell fates in Lineage And RNA RecoverY (LARRY) barcoded
hematopoietic cells. LARRY leverages unique lentiviral barcodes inserted
into 3' untranslated region (UTR) of GFP reporter that is integrated as
a single copy into parental cells. Then progeny can be clonally traced
to follow the descendent cell fates over time. LARRY has been recently
applied to in vitro differentiation of human blood and accurately
recovered expected cell fate and lineage trajectories
\citep{Weinreb2020-mv}. This dataset sampled differentiation over the
course of 2, 4, and 6 days and contained \(49,302\) cells that each
could be traced with at least one barcode. Because LARRY couples
single-cell transcriptomics and direct cell lineage tracing, this data
set provides a unique opportunity to benchmark RNA Velocity
quantitatively and to provide insights into the correct interpretation
of recovered velocity vector fields and latent cell times. To
quantitatively assess the agreement between observed and predicted
fates, we considered two metrics: 1. average cosine similarity (cs)
between vectors from the RNA Velocity vector field and the clonal
progression vector field, i.e., a vector field derived directly from the
lineage data (Sec.~\ref{sec-methods}), 2. Spearman's correlation
\((\rho)\) between cell latent (shared) time and fate potency scores
derived from Cytotrace and Cospar. Cospar, our gold standard, is a
state-of-the-art method designed explicitly for predicting fate potency
based on LARRY data, while Cytotrace leverages only transcriptomic data.
We selected cell state changes associated with monocyte and neutrophil
lineages because both lineages have the largest numbers of cells in the
scRNA sequencing data set and well-known cell fate trajectories
\citep{Petukhov2018-nr}. As expected, the Pyro -Velocity vector field
accurately predicted cell state trajectories from hematopoietic
progenitor cells into mature monocytes or neutrophils and had high
cosine similarity between the inferred vector field and clonal
progression vector field assigned by LARRY (monocytes \(cs=0.49\),
neutrophils \(cs=0.38\)). High Spearman's correlation was also seen
between shared latent time and Cospar fate potency (monocytes
\(\rho=0.60\), neutrophils \(\rho=0.74\)) or shared latent time with
Cytotrace fate potency (monocytes \(\rho=0.925\), neutrophils
\(\rho=0.925\)). By contrast, scVelo generated vector fields with
reversed order or low correlation (cosine similarity: \(-0.38\),
monocytes, cosine similarity: \(0.32\), neutrophils). Moreover, Cospar
fate potency scores were negative (\(\rho=-0.59\), monocytes) or
non-significantly correlated (\(\rho=-0.07\), neutrophils). A similar
low correlation was observed for scVelo and Cytotrace fate potency
scores (monocytes \(\rho=0.309\), neutrophils \(\rho=0.751\))
(Fig.~\ref{fig-lineage-tracing}a-b,
Supplementary Figure~\ref{suppfig-lineage-tracing}c).

Next, we evaluated the ability of the Pyro -Velocity models to recover
bi-fate or multi-fate trajectories by considering clones belonging to
monocyte and neutrophil lineages or using the entire dataset. Not
surprisingly, both methods showed their limitations, given that they are
not designed to model and capture multiple fate trajectories
(Supplementary Figure~\ref{suppfig-lineage-tracing-2}c-d). The recovered
vector fields and cell latent (shared) time show inconsistent directions
for both methods. However, Pyro -Velocity identified higher uncertainty
for some of the problematic regions. Furthermore, we discovered that
vector field and shared time estimation could be significantly improved
by extending the original kinetic model by allowing each gene a time lag
for the transcriptional activation and a baseline expression level
potentially different than 0 (Pyro -Velocity Model 2,
Sec.~\ref{sec-methods}) (Fig.~\ref{fig-lineage-tracing}c,d,
Supplementary Figure~\ref{suppfig-graphical-models}b,c). This extended
model recapitulated a vector field consistent with the clonal
progression vector field for most branches with some spurious
predictions within the intermediate cell state between stem cells and
Megakaryocytes and most terminal Neutrophils (bi-fate cosine similarity:
\(0.62\), multi-fate cosine similarity: \(0.63\)). However, the proposed
uncertainty estimation accurately flagged these subpopulations to warn
users of potential false discoveries. Notably, this extended model
significantly improved the correlation of cell shared time with the
clonal fate potency (bi-fate \(\rho=0.65\), multi-fate \(\rho=0.74\))
and also improved the correlation with Cytotrace fate potency (bi-fate
\(\rho=0.931\), multi-fate \(\rho=0.614\)). Remarkably, the shared time
per cell from Pyro -Velocity also outperformed Cytotrace for
developmental order prediction
(Supplementary Figure~\ref{suppfig-lineage-tracing}a,c). Importantly,
this extended model retained its ability to accurately identify cell
lineage trajectories and shared time for all the other datasets analyzed
above (Supplementary Figure~\ref{suppfig-model-2-result}). Taken
together, our approach using LARRY barcoded datasets established the
first quantitative benchmark of RNA velocity-based cell fate trajectory
inference and provides an exciting new framework for continued
optimization of multi-fate predictions in the future.

In conclusion, Pyro-Velocity provides the first probabilistic,
interpretable, end-to-end Bayesian inference framework for RNA Velocity
that learns cell dynamics and recovers cell fate from multimodal raw
read counts. This approach provides uncertainty estimations of cell fate
trajectories and novel visualization approaches to mitigate false
discoveries. Pyro-Velocity recovered shared time and vector fields that
are recapitulated in true developmental order and outperformed
state-of-the-art methods like Cytotrace and scVelo
(Supplementary Figure~\ref{suppfig-lineage-tracing}a).

\section{Software availability}\label{software-availability}

We developed our Pyro-Velocity models with Pyro (version 1.6.0)
probabilistic programming language and PyTorch (version 1.8.1).
Pyro-Velocity source code is available at
\url{https://github.com/pinellolab/pyrovelocity}. The version of
Pyro-Velocity (v0.1.0) used to generate the results presented in this
manuscript has been deposited on Zenodo:
\url{https://zenodo.org/record/7072876\#.Yx-2e-zMLxg}.

\section{Acknowledgments}\label{acknowledgments}

\section{Author contributions}\label{author-contributions}

\section{Competing financial interests
statement}\label{competing-financial-interests-statement}

We assume the dynamical gene expression is determined by the RNA
splicing process, and infer the unspliced and spliced gene expression
level from the differential equations proposed in velocyto
\citep{La_Manno2018-lj} and scVelo \citep{Bergen2020-pj}

\begin{align}
\frac{d u\left(\tau^{\left(k_{cg}\right)}\right)}{d \tau^{\left(k_{cg}\right)}}
  &= \alpha^{\left(k_{cg}\right)}-\beta_g u\left(\tau^{\left(k_{cg}\right)}\right), \label{eq-dudt}\\
\frac{d s\left(\tau^{\left(k_{cg}\right)}\right)}{d \tau^{\left(k_{cg}\right)}}
  &= \beta_g u\left(\tau^{\left(k_{cg}\right)}\right)-\gamma_g s\left(\tau^{\left(k_{cg}\right)}\right). \label{eq-dsdt}
\end{align}

\section{References}\label{references}

@@

\section{Supplementary Figures}\label{supplementary-figures}

@@

\caption{\label{suppfig-plot}An example figure (an empty plot)}

\renewcommand\refname{Supporting Information}

>>>>

\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else  
    % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
% Make \paragraph and \subparagraph free-standing
\makeatletter
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}{
    \@ifstar
      \xxxParagraphStar
      \xxxParagraphNoStar
  }
  \newcommand{\xxxParagraphStar}[1]{\oldparagraph*{#1}\mbox{}}
  \newcommand{\xxxParagraphNoStar}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}{
    \@ifstar
      \xxxSubParagraphStar
      \xxxSubParagraphNoStar
  }
  \newcommand{\xxxSubParagraphStar}[1]{\oldsubparagraph*{#1}\mbox{}}
  \newcommand{\xxxSubParagraphNoStar}[1]{\oldsubparagraph{#1}\mbox{}}
\fi
\makeatother

@@

\raggedbottom
\usepackage{nameref}
\usepackage{placeins}
\allowdisplaybreaks
\makeatletter
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{plain}
\@ifundefined{c@chapter}{\newfloat{suppfig}{h}{loextfig}}{\newfloat{suppfig}{h}{loextfig}[chapter]}
\floatname{suppfig}{Extended Data Fig.}
\newcommand*\listofsuppfigs{\listof{suppfig}{List of Extended Data Figures}}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother

@@

\section*{Main text}\label{sec-main}
\addcontentsline{toc}{section}{Main text}

@@

Pyro -Velocity recasts the velocity estimation problem into a latent
variable posterior probability inference
(\hyperref[sec-methods]{Methods}). The proposed model is generative and
fully Bayesian, with the different parameters considered as latent
random variables. Each parameter is assigned a prior distribution, and
the model is then conditioned on the raw data (both spliced and
unspliced molecule raw counts) to estimate posterior probability
distributions and to enable uncertainty estimation of velocity. Central
to the Pyro -Velocity model is a \emph{shared latent time} \((t)\): a
global latent variable placing each cell at a particular instant of the
modeled dynamical process. From this latent time, the expectation of
spliced and unspliced RNA levels for each gene can be computed knowing
gene-specific kinetics parameters (transcriptional \((\alpha)\),
splicing \((\beta)\), and degradation rates (\(\gamma\)) and two
gene-specific switching times \((t_{0}^{k})\), and gene-specific
transcriptional state \(k\)) and integrating the RNA velocity
differential equations analogously to what proposed in the dynamical RNA
velocity model by Bergen et al. \citep{Bergen2020-pj, Li2021-qa}.
However, we innovate the structure and formulation of the latent time
given that our model is set to jointly fit the entire raw data matrix,
not individual genes. We approach the task of estimating the posterior
distribution by restricting the search within a family of a variational
distribution approximating the true posterior, and taking advantage of
the automatic differentiation variational inference (ADVI) capabilities
of Pyro \citep{Bingham2018-id, Kucukelbir2016-bk}
(Fig.~\ref{fig-model}a, \hyperref[sec-methods]{Methods}). Notably, our
generative Bayesian model enables to sample from the posterior
distributions of the latent and dependent variables, allowing
uncertainty estimation of phase portraits, high dimensional velocity
vector field, and shared latent time. Around this model, we built
several visualizations to assess uncertainty and functions to recover
putative cell fate priming genes and to investigate how the denoised
gene expression recapitulates known cell-type annotations and their
developmental order (Fig.~\ref{fig-model}b,
\hyperref[sec-methods]{Methods}). Although several RNA velocity methods
\citep{Gayoso2024-fz, Gorin2022-il, Gu2022-rh, Cui2022-em, Gao2022-jq}
have been recently proposed, here we showcase the advantages and
innovations of Pyro -Velocity by comparing it with the current
state-of-the-art dynamical RNA velocity model proposed by Bergen et al.
\citep{Bergen2020-pj}, and implemented in scVelo.

@@

Pyro -Velocity was also able to define well-known developmental cell
hierarchies in a scRNA-seq dataset of E15.5 mouse pancreas
(Fig.~\ref{fig-model-results}b-e, \(n=3,696\) cells comprising \(8\)
cell types (Fig.~\ref{fig-model-results}b, panel 1)), identifying cell
trajectories originating from ductal progenitor cells and culminated in
the production of mature Alpha, Beta, Delta, and Epsilon cells
\citep{Bastidas-Ponce2019-lf}. Though the scVelo and Pyro -Velocity
models have qualitatively comparable vector fields, the results
generated by scVelo were greatly influenced by gene selection,
neighboring cell number for gene expression smoothing, and cannot
provide confidence assessment of lineage directionality
(Extended Data Fig.~\ref{extfig-model-results}a). In addition,
Pyro -Velocity uncertainty visualizations uncovered that differentiated
cells had a larger vector field uncertainty than stem and progenitor
cells, as would be expected for terminally differentiated cell states
where progression and lineage trajectories would not be anticipated.
Pyro -Velocity also recovered a developmental order in close agreement
with what was estimated by Cytotrace (Spearman's
\(r=0.95, \; \text{p-value} <
10^{-10}\) between latent time and developmental
order)\citep{Gulati2020-xq} - a state-of-the-art orthogonal method used
to predict cell differentiation based on the number of expressed genes
per cell.

We next used Pyro -Velocity to unbiasedly uncover potential cell fate
determinant markers genes within the E15.5 mouse pancreas. To this end,
we first selected the top 300 genes with the best goodness of fit
(negative mean absolute errors across cells). Next, we ranked them based
on the positive (or negative) correlation between inferred shared time
and corresponding denoised spliced expression per cell
(Fig.~\ref{fig-model-results}d, Methods). This procedure successfully
recovered key developmental genes that are activated (or repressed)
during endocrinogenesis (Fig.~\ref{fig-model-results}d-e,
Extended Data Fig.~\ref{extfig-model-results}). Among the top positively
correlated genes identified by Pyro -Velocity, endocrine Beta cells
expressed insulin 1 and 2 (Ins1, Ins2) and Nnat, which is a key
regulator Beta cell insulin generation \citep{Millership2018-vd} and
Ppp1r1a which is a marker of Beta cells \citep{Jiang2013-pt}. The
ranking procedure proposed by scVelo failed to identify Ins1, Ins2,
Nnat, and Ppp1r1a as lineage-restricted genes in their top 50 ranked
gene list. The inability of scVelo to identify these key developmental
genes could be likely explained by the kNN smoothing of read count and
filtering steps (Supplementary Table 1, 2). Visual inspection of these
four top genes by Pyro -Velocity revealed a single activation phase
during cell type transitions based on phase portrait curves. We also
propose their visualization through ``rainbow plots'\,' to illustrate
denoised single gene spliced expression across shared time and to
investigate the inferred cell order based on known annotations. Further,
UMAP embedding of cells colored based on denoised gene expression
confirmed the cell fate markers' smooth and progressive activation (or
repression) towards endocrine cells (Fig.~\ref{fig-model-results}e,
Extended Data Fig.~\ref{extfig-model-results}).

Ground truth velocity is not a quantity we can easily assess. Therefore,
no robust and quantitative benchmarking of RNA velocity methods has been
proposed. However, recent single-cell assays allow recording lineage
relationships and can be used to assess the ability of different methods
in recapitulating cell fate progression. Therefore, we next
systematically compared scVelo and Pyro -Velocity ability to recover
single cell fates in Lineage And RNA RecoverY (LARRY) barcoded
hematopoietic cells. LARRY leverages unique lentiviral barcodes inserted
into 3' untranslated region (UTR) of GFP reporter that is integrated as
a single copy into parental cells. Then progeny can be clonally traced
to follow the descendent cell fates over time. LARRY has been recently
applied to in vitro differentiation of human blood and accurately
recovered expected cell fate and lineage trajectories
\citep{Weinreb2020-mv}. This dataset sampled differentiation over the
course of 2, 4, and 6 days and contained \(49,302\) cells that each
could be traced with at least one barcode. Because LARRY couples
single-cell transcriptomics and direct cell lineage tracing, this data
set provides a unique opportunity to benchmark RNA Velocity
quantitatively and to provide insights into the correct interpretation
of recovered velocity vector fields and latent cell times. To
quantitatively assess the agreement between observed and predicted
fates, we considered two metrics: 1. average cosine similarity (cs)
between vectors from the RNA Velocity vector field and the clonal
progression vector field, i.e., a vector field derived directly from the
lineage data (\hyperref[sec-methods]{Methods}), 2. Spearman's
correlation \((\rho)\) between cell latent (shared) time and fate
potency scores derived from Cytotrace and Cospar. Cospar, our gold
standard, is a state-of-the-art method designed explicitly for
predicting fate potency based on LARRY data, while Cytotrace leverages
only transcriptomic data. We selected cell state changes associated with
monocyte and neutrophil lineages because both lineages have the largest
numbers of cells in the scRNA sequencing data set and well-known cell
fate trajectories \citep{Petukhov2018-nr}. As expected, the
Pyro -Velocity vector field accurately predicted cell state trajectories
from hematopoietic progenitor cells into mature monocytes or neutrophils
and had high cosine similarity between the inferred vector field and
clonal progression vector field assigned by LARRY (monocytes
\(cs=0.49\), neutrophils \(cs=0.38\)). High Spearman's correlation was
also seen between shared latent time and Cospar fate potency (monocytes
\(\rho=0.60\), neutrophils \(\rho=0.74\)) or shared latent time with
Cytotrace fate potency (monocytes \(\rho=0.925\), neutrophils
\(\rho=0.925\)). By contrast, scVelo generated vector fields with
reversed order or low correlation (cosine similarity: \(-0.38\),
monocytes, cosine similarity: \(0.32\), neutrophils). Moreover, Cospar
fate potency scores were negative (\(\rho=-0.59\), monocytes) or
non-significantly correlated (\(\rho=-0.07\), neutrophils). A similar
low correlation was observed for scVelo and Cytotrace fate potency
scores (monocytes \(\rho=0.309\), neutrophils \(\rho=0.751\))
(Fig.~\ref{fig-lineage-tracing}a-b,
Extended Data Fig.~\ref{extfig-lineage-tracing}c).

Next, we evaluated the ability of the Pyro -Velocity models to recover
bi-fate or multi-fate trajectories by considering clones belonging to
monocyte and neutrophil lineages or using the entire dataset. Not
surprisingly, both methods showed their limitations, given that they are
not designed to model and capture multiple fate trajectories
(Extended Data Fig.~\ref{extfig-lineage-tracing-2}c-d). The recovered
vector fields and cell latent (shared) time show inconsistent directions
for both methods. However, Pyro -Velocity identified higher uncertainty
for some of the problematic regions. Furthermore, we discovered that
vector field and shared time estimation could be significantly improved
by extending the original kinetic model by allowing each gene a time lag
for the transcriptional activation and a baseline expression level
potentially different than 0 (Pyro -Velocity Model 2,
\hyperref[sec-methods]{Methods}) (Fig.~\ref{fig-lineage-tracing}c,d,
Extended Data Fig.~\ref{extfig-graphical-models}b,c). This extended
model recapitulated a vector field consistent with the clonal
progression vector field for most branches with some spurious
predictions within the intermediate cell state between stem cells and
Megakaryocytes and most terminal Neutrophils (bi-fate cosine similarity:
\(0.62\), multi-fate cosine similarity: \(0.63\)). However, the proposed
uncertainty estimation accurately flagged these subpopulations to warn
users of potential false discoveries. Notably, this extended model
significantly improved the correlation of cell shared time with the
clonal fate potency (bi-fate \(\rho=0.65\), multi-fate \(\rho=0.74\))
and also improved the correlation with Cytotrace fate potency (bi-fate
\(\rho=0.931\), multi-fate \(\rho=0.614\)). Remarkably, the shared time
per cell from Pyro -Velocity also outperformed Cytotrace for
developmental order prediction
(Extended Data Fig.~\ref{extfig-lineage-tracing}a,c). Importantly, this
extended model retained its ability to accurately identify cell lineage
trajectories and shared time for all the other datasets analyzed above
(Extended Data Fig.~\ref{extfig-model-2-result}). Taken together, our
approach using LARRY barcoded datasets established the first
quantitative benchmark of RNA velocity-based cell fate trajectory
inference and provides an exciting new framework for continued
optimization of multi-fate predictions in the future.

In conclusion, Pyro -Velocity provides the first probabilistic,
interpretable, end-to-end Bayesian inference framework for RNA Velocity
that learns cell dynamics and recovers cell fate from multimodal raw
read counts. This approach provides uncertainty estimations of cell fate
trajectories and novel visualization approaches to mitigate false
discoveries. Pyro -Velocity recovered shared time and vector fields that
are recapitulated in true developmental order and outperformed
state-of-the-art methods like Cytotrace and scVelo
(Extended Data Fig.~\ref{extfig-lineage-tracing}a).

\section*{References}\label{references}
\addcontentsline{toc}{section}{References}

@@

\twocolumn

We assume the dynamical gene expression is determined by the RNA
splicing process, and infer the unspliced and spliced gene expression
level from the differential equations proposed in velocyto
\citep{La_Manno2018-lj} and scVelo \citep{Bergen2020-pj} \begin{align}
\frac{d u\left(\tau^{\left(k_{cg}\right)}\right)}{d \tau^{\left(k_{cg}\right)}}
  &= \alpha^{\left(k_{cg}\right)}-\beta_g u\left(\tau^{\left(k_{cg}\right)}\right), \label{eq-dudt}\\
\frac{d s\left(\tau^{\left(k_{cg}\right)}\right)}{d \tau^{\left(k_{cg}\right)}}
  &= \beta_g u\left(\tau^{\left(k_{cg}\right)}\right)-\gamma_g s\left(\tau^{\left(k_{cg}\right)}\right). \label{eq-dsdt}
\end{align}

\subsection{Model evaluation and selection}\label{sec-model-selection}

Pyro -Velocity provides two model formulations as described above. Users
can compare their predictive performance to decide which of the two
models to use for a given dataset. Namely, the mean absolute errors of
posterior samples prediction of spliced and unspliced raw read counts
\(x_{icg}\) from its observation \(\tilde{X}_{cg}\) per gene and cell
combinations \begin{align*}
\frac{1}{N_i N_c N_g} \sum_{i=1}^{N_i} \sum_{c=1}^{N_c} \sum_{g=1}^{N_g} |x_{icg} - \tilde{X}_{cg}|,
\end{align*} where \(i\) is the posterior sample index, \(N_i\) is the
posterior sample number, \(N_c\) is the cell number, \(N_g\) is the gene
number. Then it is possible to evaluate the cell fate inference
performance of the models based on two metrics:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  the consistency between the velocity vector field and the \emph{clonal
  progression vector field} using grid-based cosine similarity; 2.
  Spearman's correlation of velocity-based shared (latent) time with
  Cospar fate potency and the silver standard Cytotrace score. Both
  metrics are implemented in scipy.
\end{enumerate}

\section{Code availability}\label{code-availability}

We developed our Pyro -Velocity models with Pyro (version 1.6.0)
probabilistic programming language and PyTorch (version 1.8.1).
Pyro -Velocity source code is available at
\url{https://github.com/pinellolab/pyrovelocity}. The version of
Pyro -Velocity (v0.1.0) used to generate the results presented in this
manuscript has been deposited on Zenodo:
\url{https://zenodo.org/record/7072876\#.Yx-2e-zMLxg}.

\section*{Acknowledgments}\label{acknowledgments}
\addcontentsline{toc}{section}{Acknowledgments}

\section*{Author contributions}\label{author-contributions}
\addcontentsline{toc}{section}{Author contributions}

\section*{Competing financial interests
statement}\label{competing-financial-interests-statement}
\addcontentsline{toc}{section}{Competing financial interests statement}

\section{Extended Data Figures}\label{extended-data-figures}

@@
